\MakeUppercase{Graph Product Structure for Non-Minor-Closed Classes}

### 2022-03-24 k-planar.tex
Got rid of reference to last section
Formatting
Pffft

\begin{table}[h]
  \begin{center}
    \begin{tabular}{lccl}
      \textbf{Graph class} & \boldmath $\tw(H)$ & \boldmath $\ell$ & \textbf{Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      planar & $6^*$ & $1$ & \cite{UWY} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW} \\
%      genus $g$ & $2g+6$ & 1 & \cite{UWY} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\[1ex]
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-NN & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline
      \\[.2ex]
      \multicolumn{4}{l}{$^*$ the bound also applies to the simple treewidth of $H$.}
    \end{tabular}
  \end{center}
  \caption{Known product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for string graphs, powers of bounded-degree graphs, map graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure and the consequences of the current work.
  \item \cref{WrappingUp} concludes with an open problem.
\end{compactitem}

@@

	The proof is by induction on the number of inner faces of $N'$.  A particular easy case occurs when $V(N')=V(F)$, for which we can simply take
	\note{DW}{Suggested revision, ``In the base case, $N'$ has only one inner face, and we can simply take ...} \note{PM}{I debated what to put here since, even when $N'$ has only one face, the rules below work for picking $\tau$.  The \emph{real} base case occurs when $N'$ has zero faces, but that's not something that ever happens since we only ever apply induction on the inner faces of $M$ (except $\tau$).  I settled on this ``base case'' since it's clear that, in this case, the partition $\mathcal{P}$ is already decided.}
	$\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iv):
	\begin{compactenum}[(i)]
		\item By definition, each of $P_1,\ldots,P_k$ is a part of $\mathcal{P}$.
		\item Assumption 4(b) ensures, for each $x\in V(H)$, the existence of $X_x$ and $Y_x$ such that $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each integer $j\ge 0$.
		\item The trivial tree decomposition of $H$ that has a single bag containing $\{x_1,\ldots,x_k\}$ is $3$-simple and, if $k=3$, has exactly one bag that contains $\{x_1,\ldots,x_3\}$.
	\end{compactenum}

@@

Since all the product structure theorems presented thus far upper bound the treewidth of $H$,
\cref{qn} immediately implies bounds on the queue number of all graphs in these classes.  An improvement can be made in the cases where $H$ is planar and has treewidth $3$, as follows.
\citet{ABGKP20} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1_planar_product,dMapProduct} has queue-number at most $5$, which with \cref{qn} implies:

\begin{cor}
\label{1PlanarQueue}
Every 1-planar graph has queue-number at most 
$3 \times 7 \times 5 + \floor{\tfrac{3}{2} \times 7} = 115$.
\end{cor}

\begin{cor}
\label{dMapQueue}
Every $d$-map graph has queue-number at most 
$\floor{ \frac{33}{2} (d+3\floor{d/2}-3) }$.
\note{DW}{I have updated this bound using the better qn bound for planar 3-trees. Note that with $d=3$ we get the original bound of 49.}
\end{cor}

% \cref{qn,gkPlanarProduct} imply that $(g,k)$-planar graphs have queue-number at most $g 2^{O(k^3)}$.
% \note{DW}{I have changed $g 2^{O(k^4)}$ to $g 2^{O(k^3)}$, okay?}
% \note{OM}{Ok}
% \note{DW}{I have changed $g>2^{k^3}$ to $g>2^{k^2}$, okay?}.
% \note{PM}{Ok}

The following corollary summarizes the results for other graph classes.
\begin{cor}\label{q_cor}
  The following bounds hold for the queue number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $2^{O(k^3)}$
    \item $(g,k)$-planar: $\max\{1,g\}\cdot 2^{O(k^3)}$
    \item $(g,\delta)$-string: $\max\{1,g\}\cdot 2^{O(\delta^3)}$
    \item $1$-planar: $115$ \note{DW}{delete?}
    \item $d$-map: $24 (d+3\floor{d/2}-3) \,  - \ceil{\tfrac{3}{2} (d+3\floor{d/2}-3)}$ \note{DW}{delete?}
    \item $(g,d)$-map: $\max\{1,g\}\cdot 2^{O(d)}$
    \item $k$-NN: $2^{O(k^6)}$
    % \item $(g,d)$-map graph
    % \item
  \end{compactenum}
\end{cor}

@@

For planar graphs of treewidth at most $3$ an $O(p^2\log p)$ upper bound is known:
\note{DW}{I prefer ``a $O(p^2\log p)$'', pronounced ``a big-O...''}

@@

As with nonrepetitive chromatic number, our results also imply bounds on the $p$-centered chromatic numer of $(g,\delta)$-string graphs and $k$-nearest-neighbour graphs, but these bounds are weaker than the bounds implied by the fact that these graphs have bounded maximum degree \cite{DFMS21}.

>>>>

\begin{table}[h]
  \centering{
    \begin{tabular}{lccl}
      \textbf{Graph class} & \boldmath $\tw(H)$ & \boldmath $\ell$ & \textbf{Reference} \\ \hline
      planar & $3^*$ & $3$ & \cite{DJMMUW20} \\
      planar & $6^*$ & $1$ & \cite{UWY} \\
      genus $g$ & $3^*$ & $\max\{2g,3\}$ & \cite{DHHW} \\
%      genus $g$ & $2g+6$ & 1 & \cite{UWY} \\
      apex-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\[1ex]
      % B-minor-free & $O(1)$ & $O(1)$ & \cite{DJMMUW20} \\
      $k$-planar & $\binom{k+4}{3}-1$ & $18k^2 + 48k + 30$ & \cref{kPlanarProduct} \\
      $(g,k)$-planar & $\binom{k+4}{3}-1$ & $\max\{2g,3\}(6k^2+16k+10)$ & \cref{gkPlanarProduct} \\
      $(g,\delta)$-string & $\binom{\delta+4}{3}-1$ & $\max\{2g,3\}(\delta^4 + 4\delta^3 + 9\delta^2 + 10\delta+4)$ & \cref{StringProduct} \\
      $k$-NN & $O(k^6)$ & $O(k^4)$ & \cref{nn_product_structure} \\
      $d$-framed & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{d_framed_product_stucture} \\
      $1$-planar & $3^*$ & $7$ & \cref{1_planar_product} \\
      $d$-map & $3^*$ & $d+ 3\floor{d/2} - 3$ & \cref{dMapProduct} \\
      $(g,d)$-map & $9$ & $\max\{2g,3\}\,(7d^2 -21d)$ & \cref{gdMapProduct} \\ \hline 
      \multicolumn{4}{l}{$^*$ these bounds also applies to the simple treewidth of $H$.}
    \end{tabular}
  }
  \caption{Known product structure theorems (of the form $G\subseteq H\boxtimes P\boxtimes K_\ell$).}
  \label{summary_table}
\end{table}

The remainder of this paper is organized as follows:
\begin{compactitem}
  \item In \cref{Structure} we prove our main result for shortcut systems (\cref{ShortcutProduct}), and its optimisations for $k$-planar graphs and their genus-$g$ generalisation (\cref{kPlanarProduct,gkPlanarProduct}).
  \item \Cref{examples} uses \cref{ShortcutProduct,kPlanarProduct,gkPlanarProduct} to derive product structure theorems for string graphs, powers of bounded-degree graphs, map graphs, and $k$-nearest neighbour graphs.
  \item \Cref{FramedSection} proves a product structure theorem for $d$-framed multigraphs in which the graph $H$ has treewidth $3$ and is planar.  Using this result, we derive improved product structure theorems for $1$-planar graphs and planar map graphs, again with a planar graph $H$.
  \item \Cref{Applications} quickly surveys applications of product structure and the consequences of the current work.
\end{compactitem}

@@

	The proof is by induction on the number of inner faces of $N'$.  A particular easy case occurs when $V(N')=V(F)$, for which we can simply take
	\note{DW}{Suggested revision, ``In the base case, $N'$ has only one inner face, and we can simply take ...} \note{PM}{I debated what to put here since, even when $N'$ has only one face, the rules below work for picking $\tau$.  The \emph{real} base case occurs when $N'$ has zero faces, but that's not something that ever happens since we only ever apply induction on the inner faces of $M$ (except $\tau$).  I settled on this ``base case'' since it's clear that, in this case, the partition $\mathcal{P}$ is already decided.}
  \note{PM}{The longwinded version is that induction is on the pair $(x,y)$ where $x=|V(N-F)|$ and $y$ is the number of inner faces of $N'$.  The base cases are $(0,y)$ for any $y$.}
	$\PP:=\{P_1,\ldots,P_k\}$ and verify that the preconditions of the lemma (the `setup') ensure that $\PP$ satisfies the requirements (i)--(iii):
	\begin{compactenum}[(i)]
		\item By definition, each of $P_1,\ldots,P_k$ is a part of $\mathcal{P}$.
		\item Assumption 4(b) ensures, for each $x\in V(H)$, the existence of $X_x$ and $Y_x$ such that $|X_x|\le d-3$ and $|Y_x\cap L_j|\le 3$ for each integer $j\ge 0$.
		\item The trivial tree decomposition of $H$ that has a single bag containing $\{x_1,\ldots,x_k\}$ is $3$-simple and, if $k=3$, has exactly one bag that contains $\{x_1,\ldots,x_3\}$.
	\end{compactenum}

@@

Since all the product structure theorems presented thus far upper bound the treewidth of $H$,
\cref{qn} immediately implies bounds on the queue number of all graphs in these classes.  An improvement can be made for $1$-planar and $d$-map graphs since, in the cases $H$ is planar and has treewidth $3$.
\citet{ABGKP20} proved that every planar graph with treewidth at most $3$ has queue-number at most $5$. Thus the graph $H$ in \cref{1_planar_product,dMapProduct} has queue-number at most $5$.
% , which with \cref{qn} implies:
% 
% \begin{cor}
% \label{1PlanarQueue}
% Every 1-planar graph has queue-number at most 
% $3 \times 7 \times 5 + \floor{\tfrac{3}{2} \times 7} = 115$.
% \end{cor}
% 
% \begin{cor}
% \label{dMapQueue}
% Every $d$-map graph has queue-number at most 
% $\floor{ \frac{33}{2} (d+3\floor{d/2}-3) }$.
% \note{DW}{I have updated this bound using the better qn bound for planar 3-trees. Note that with $d=3$ we get the original bound of 49.}
% \end{cor}
% 
% 
% 
% \cref{qn,gkPlanarProduct} imply that $(g,k)$-planar graphs have queue-number at most $g 2^{O(k^3)}$.
% \note{DW}{I have changed $g 2^{O(k^4)}$ to $g 2^{O(k^3)}$, okay?}
% \note{OM}{Ok}
% \note{DW}{I have changed $g>2^{k^3}$ to $g>2^{k^2}$, okay?}.
% \note{PM}{Ok}
The following corollary summarizes all of these results.
\begin{cor}\label{q_cor}
  The following bounds hold for the queue number of any graph from each of the following classes:
  \begin{compactenum}
    \item $k$-planar: $2^{O(k^3)}$
    \item $(g,k)$-planar: $\max\{1,g\}\cdot 2^{O(k^3)}$
    \item $(g,\delta)$-string: $\max\{1,g\}\cdot 2^{O(\delta^3)}$
    \item $1$-planar: $115$ 
    \item $d$-map: $\floor{ \frac{33}{2} (d+3\floor{d/2}-3) }$ 
    \item $(g,d)$-map: $\max\{1,g\}\cdot 2^{O(d)}$
    \item $k$-NN: $2^{O(k^6)}$
    % \item $(g,d)$-map graph
    % \item
  \end{compactenum}
\end{cor}
\note{PM}{For uniformity, I got rid of the two corollaries on $d$-map and $1$-planar graphs and kept them in \cref{q_cor}.  The text is still commented out in the source in case someone feels strongly about keeping them.}

@@

For planar graphs of treewidth at most $3$ a $O(p^2\log p)$ upper bound is known:
% \note{DW}{I prefer ``a $O(p^2\log p)$'', pronounced ``a big-O...''}

@@

As with nonrepetitive chromatic number, our results also imply bounds on the $p$-centered chromatic numer of $(g,\delta)$-string graphs, powers of bounded-degree graphs, and $k$-nearest-neighbour graphs, but these bounds are weaker than the bounds implied by the fact that these graphs have bounded maximum degree \cite{DFMS21}.
